---
title: "Mill River PFAS: SSN object development"
author: "Raul Vera and Matthew Fuller"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    fig_caption: yes
    toc: yes
    toc_depth: 4
  html_document:
    toc: yes
    toc_depth: '4'
    df_print: paged
    code_folding: hide 
---


__________

This document describes the process and code used to explore PFAS concentrations in the Mill River watershed near Amherst, MA using spatial stream network (SSN) models.  

Specifically, we build the SSN object in this script and it takes less than 5 minutes to run all the code.  


#  1.  Set Up Analysis  

```{r setup, message=FALSE, warning=FALSE}
library(viridis)
library(stringr)
library(knitr)
library(pander)
library(tidyverse)
library(sf)
library(SSNbler)
library(SSN2)
panderOptions('digits',8)

# set tidy r code not to run more than 100 characters wide
opts_chunk$set(tidy.opts=list(width.cutoff=100),tidy=FALSE) 

# hide all warning messages in document
knitr::opts_chunk$set(message = FALSE)
```

#  2.  Read in data  

## Network data  

```{r}
covariate_data <- readRDS("../data/summary_basin_covariates.rds") |>
  left_join(readRDS("../data/summary_rca_covariates.rds"), by = "reach_id")

flines  <- st_read(dsn = "../data/shp/MillRiver_NHDPlusHR_EPSG5070_trace.shp") |>
  left_join(covariate_data, by = "reach_id")

flowline <- st_transform(flines, crs =5070) |> st_zm() 
```


##  Observation sites  

```{r}
#bringing in the results from my PFAS analysis
PFAS_Spatial_March_2026 <- 
  read_csv("https://raw.githubusercontent.com/rvera177/MillRiver_PFAS/refs/heads/main/data/Spatial_3_PFASresults.csv")

#giving a simpler name
S3 <- PFAS_Spatial_March_2026 

S3 <- S3 %>% #removing NA's in lat and long. Shouldn't be any
  filter(!is.na(Latitude), !is.na(Longitude))

#this changes the name of some of the headers for easier manipulation later
clean_names <- function(x) {
  x |>
    gsub(" ", "_", x = _) |>
    gsub("[-:]", ".", x = _) |>
    make.names() |>
    gsub("^X(\\d)", "\\1", x = _)
}

colnames(S3) <- clean_names(colnames(S3))
names(S3)
#Naming functional head groups that I can add together later
pfas_groups <- list(
  PFCA = c("PFBA", "PFPeA", "PFHxA", "PFHpA", "PFOA",
           "PFNA", "PFDA", "PFUnA", "PFDoA", "PFTrDA", "PFTeDA"),
  PFSA = c("PFBS", "PFPeS", "PFHxS", "PFHpS", "PFOS",
           "PFNS", "PFDS", "PFDoS"),
  FTSA = c("4.2FTS", "6.2FTS", "8.2FTS"),
  PFOSA = c("PFOSA", "NMeFOSA", "NEtFOSA"),
  FOSAA = c("NMeFOSAA", "NEtFOSAA"),
  PFECA = c("HFPO.DA", "ADONA", "PFMPA", "PFMBA", "NFDHA"),
  PFESA = c("9Cl.PF3ONS", "11Cl.PF3OUdS", "PFEESA"),
  FTCA = c("3.3_FTCA", "5.3_FTCA", "7.3_FTCA"))

#create new columns in S3 with the sum of corresponding species in the family
S3 <- S3 %>%
  bind_cols(
    lapply(names(pfas_groups), function(fam_name) {
      # Get the column names in S3 for this family
      cols <- paste0(pfas_groups[[fam_name]], "_Results")
      cols_existing <- intersect(cols, names(S3))  # only existing columns
      
      tibble(
        !!fam_name := rowSums(S3[cols_existing], na.rm = TRUE)
      )
    })
  )

S3 <- S3 %>%
  mutate(OBSPRED_ID = row_number())

# use this geospatially snapped point file to join to the PFAS data
obs_nhdhr <- st_read(dsn = "../data/shp/obs_sites.shp")

# Join geospatial data with the PFAS data
S3_Cat <-  left_join(obs_nhdhr, S3, by = c("site_id"="Name")) |>
  left_join(covariate_data, by = "reach_id")

S3_sf <- st_transform(S3_Cat, st_crs(flines)) |>
  mutate(ssnid = row_number()) 

obs <- st_transform(S3_sf, crs =5070) |> st_zm() 
```


## Predictions sites  

These are placed at the center of each flowline, but will be represented by data sets aggregated/summarized for the downstream pour point of the reach/flowline. Since the NHDPlus high-resolution network has relatively short reach segments (mean length of ##m) this shouldn't be problematic or poorly representative of the reach.  

```{r}
pred <- flines |> 
  st_zm() |>                  # drop Z and M dimensions if they exist
  st_centroid() |>            # centroid works for LINESTRING
  st_cast("POINT") |>         # ensures geometry is POINT
  mutate(ssnid = reach_id + 10000) |>
  select(ssnid, reach_id, NHDPlusIDt, 
         ends_with(c("_km2","_bas","_rca")),
         lengthkm, areasqkm, totdasqkm, geometry) |>
  st_transform(pred_sites, crs =5070) |> st_zm() 
```


#  3.  Build the SSN object  

Directory to hold the files during SSN object development.  

```{r}
ssnobj_devdir <- "../ssnobj_dev/"
```

##  3.1. Develop the Landscape network (LSN) 

```{r}
flowlines_2 <- lines_to_lsn(streams = flowline, 
                           lsn_path = ssnobj_devdir,
                           check_topology = TRUE,
                           snap_tolerance = 0,
                           topo_tolerance = 0,
                           remove_ZM = TRUE,
                           overwrite = TRUE)
```

##  3.2. Incorporate sites  

###  3.2.1. Observation sites  

```{r}
obs <- sites_to_lsn(sites = obs,
                    edges = flowlines_2,
                    lsn_path = ssnobj_devdir,
                    file_name = "obs",
                    snap_tolerance = 5,
                    save_local = TRUE,
                    overwrite = TRUE)
```

###  3.2.1. Prediction sites  

```{r}
preds <- sites_to_lsn(sites = pred,
                      edges = flowlines_2,
                      lsn_path = ssnobj_devdir,
                      file_name = "preds",
                      snap_tolerance = 250,
                      save_local = TRUE,
                      overwrite = TRUE)
```

##  3.3.  Calculate upstream distances for edges and sites  

```{r}
edges <- updist_edges(edges = flowlines_2,
                      save_local = TRUE,
                      lsn_path = ssnobj_devdir,
                      calc_length = TRUE)
```


```{r}
site.list <- updist_sites(sites = list(obs = obs, 
                                       preds = preds),
                          edges = edges,
                          length_col = "Length",
                          save_local = TRUE,
                          lsn_path = ssnobj_devdir)
```


##  3.4. Calculate additive function values  


```{r}
edges <- afv_edges(edges = edges,
                   infl_col = "basin_area_km2",
                   segpi_col = "areaPI",
                   afv_col = "afvArea",
                   lsn_path = ssnobj_devdir)

site.list <- afv_sites(sites = site.list,
                       edges = edges,
                       afv_col = "afvArea",
                       save_local = TRUE,
                       lsn_path = ssnobj_devdir)
```


#  4. Assemble the pieces into an SSN object  

```{r}
PFAS_ssn <- ssn_assemble(edges = edges,
                         lsn_path = ssnobj_devdir,
                         obs_sites = site.list$obs,
                         preds_list = site.list[c("preds")],
                         ssn_path = "../ssnobj/PFAS_NHDHR.ssn",
                         import = FALSE,
                         check = TRUE, 
                         afv_col = "afvArea",
                         overwrite = TRUE,
                         verbose = TRUE)
```




